---
title: "Targeted Sequencing"
author: Elgin Akin
date: 2026-04-17
format: html
---

This is the most accurate count - starting from the ms metadata table!

# Microsattelite

## Import data

Last data import: 2024-09-12
Includes Ghana samples

```{r}
library(tidyverse)
library(ape)
library(adegenet)
library(poppr)
library(pegas)
library(igraph)

theme_set(theme_minimal(base_family = "helvetica"))

# old 
#region_colors <- c( 
#                  "Thailand" = "#003A59", 
#                  "Cambodia" = "#003A59", 
#                  "Columbia" = "#416322FF", 
#                  "Ghana" = "#A1378B",
#                  "Kenya" = "#FF5800"
#                 )

# updated colors for clarity
region_colors <- c( 
                  "Thailand" = "#EE4C97FF", 
                  "Cambodia" = "#EE4C97FF", 
                  "Colombia" = "#20854EFF", 
                  "Ghana" = "#0072B5FF",
                  "Kenya" = "#FF5800"
                 )



clone="#54bebe" 
original="#c80064"
```


```{r}
# Load the CSV file
msdf = read_csv('data/20240726_ms_data.csv')

msdf <- msdf |> 
  mutate(across(where(is.character), ~ na_if(., "RERUN")))

# metadata: Create 'country' and 'passage_history' columns with the specified logic from the master dataframe
msdf_metadata <- msdf |> 
  mutate(
    # Populate country based on the Sample column and reference names
    country = case_when(
      str_detect(str_to_lower(Sample), "3d7|7g8|fvo|nf54|dd2") ~ "reference",  # Case-insensitive check for reference names
      str_starts(Sample, "K") ~ "Kenya",
      str_starts(Sample, "G") ~ "Ghana",
      str_starts(Sample, "C") ~ "Colombia",
      str_starts(Sample, "T") ~ "Thailand",
      TRUE ~ "unknown"  # Default case
    ),
    
    # Populate passage_history based on the Sample naming convention
    passage_history = case_when(
      str_detect(str_to_lower(Sample), "cloned") ~ "clone",  # Case-insensitive check for "Cloned"
      str_detect(str_to_lower(Sample), "original vial") ~ "original",  # Case-insensitive check for "Original vial"
      str_detect(Sample, "-") ~ "clone",  # Contains a dash
      !str_detect(Sample, "-|cloned|original vial") ~ "original",  # No dash, additional strings
      TRUE ~ "unknown"  # Default case
    )
  )

# filter metadata: 
msdf_filter = msdf |> 
  drop_na(Sample) |>  # Drop NA values in Sample column
  filter(`Poly-A` != "POLYCLONAL") |>  # Drop rows where Poly-A is POLYCLONAL
  select(-`AMA1 Sequenced`, -`MSP1 Sequenced`, -`MSP2 Sequenced`, -`CSP Fullgene Sequenced`, -`Additional Notes`)  # Drop specified columns

# intersect metadata to be with filtered dataset
msdf_metadata_filtered = inner_join(msdf_metadata, msdf_filter, by = "Sample") #|> select(all_of(colnames(msdf_metadata)))  # Keep only the columns in msdf_metadata

# Add the country column to msdf_filter from msdf_metadata_filtered since adegenet and poppr can inpute a group assignment 
msdf_filter <- msdf_filter |> 
  left_join(select(msdf_metadata_filtered, Sample, country), by = "Sample")

# final data frames

msdf_metadata # use for general statistics 
msdf_metadata_filtered # use for filtered msdf
msdf_filter # filtered msdf

```

## Metadata Stats

-   sample by country distribution.
-   ? monoclonal to polyclonal as detected by ms
-   upset plot microsattelite intersect with sequences

## Microsatelite Processing

-   convert filtered ms df to geneid object
-   general stats with poppr
-   breakdown of "clones" by country origin - 4 donut charts

## Import Microsattelite Data

```{r}
pfms <- df2genind(
  X=msdf_filter[,c(2:13)], 
  ind.names= msdf_filter$Sample,
  pop=msdf_filter$country,
  NA.char="0",
  sep = ",",
  type = "codom", 
  ploidy = 1
)
```

## poppr population diversity statistics

```{r}
# Load necessary libraries
library(gt)
library(ggplot2)
library(dplyr)

# Your poppr function call
df_poppr = poppr(
  pfms,
  total = TRUE,
  sublist = "ALL",
  exclude = NULL,
  blacklist = NULL,
  sample = 0,
  method = 1,
  missing = "ignore",
  cutoff = 0.05,
  quiet = FALSE,
  clonecorrect = FALSE,
  strata = 1,
  keep = 1,
  plot = TRUE,
  hist = TRUE,
  index = "rbarD",
  minsamp = 10,
  legend = FALSE
)

# Print the poppr output
print(df_poppr)

# --- Plot the poppr output table using gt

# Create a gt table, filtering out the 'File', 'MLG', 'eMLG', and 'SE' columns
gt_table <- df_poppr %>%
  select(-c(File, MLG, eMLG, SE)) %>%  # Remove the 'File', 'MLG', 'eMLG', and 'SE' columns
  gt() %>%
  tab_header(
    title = "Microsatellite Diversity by Population "
  ) %>%
  cols_label(
    Pop = "Population",
    N = "Sample Size (N)",
    H = "Shannon-Weiner Diversity Index (H)",
    G = "Stoddard-Taylor's Index (G)",
    lambda = "Simpson's Index (λ)",
    E.5 = "Evenness (E.5)",
    Hexp = "Expected Heterozygosity (Hexp)",
    Ia = "Index of Association (Ia)",
    rbarD = "Standardized Index of Association (rbarD)"
  ) %>%
  fmt_number(
    columns = vars(N, H, G, lambda, E.5, Hexp, Ia, rbarD),
    decimals = 2
  ) %>%
  tab_options(
    table.width = pct(100),  # Set table width to 100%
    table.border.top.color = "black",
    table.border.bottom.color = "black",
    column_labels.border.top.color = "black",
    column_labels.border.bottom.color = "black"
  ) %>%
  cols_align(
    align = "center",
    columns = everything()
  )

# Print the gt table
print(gt_table)

# Save the gt table to an HTML file for larger viewing
gtsave(gt_table, "population_statistics.html")


```

## Locus Specific Descriptors

```{r}
library(gt)
library(dplyr)

# --- Plot population totals
region_colors <- c( 
                  "Thailand" = "#EE4C97FF", 
                  "Cambodia" = "#EE4C97FF", 
                  "Colombia" = "#20854EFF", 
                  "Ghana" = "#0072B5FF",
                  "Kenya" = "#FF5800"
                 )

# Create locus tables for each population
locus_table_kenya <- locus_table(
  pfms,
  population = "Kenya",
  index = "simpson",
  lev = "allele",
  information = TRUE
)

locus_table_ghana <- locus_table(
  pfms,
  population = "Ghana",
  index = "simpson",
  lev = "allele",
  information = TRUE
)

locus_table_thailand <- locus_table(
  pfms,
  population = "Thailand",
  index = "simpson",
  lev = "allele",
  information = TRUE
)

locus_table_colombia <- locus_table(
  pfms,
  population = "Colombia",
  index = "simpson",
  lev = "allele",
  information = TRUE
)

# ------------- Plot 

# Assuming the locus_table functions have already been executed for each population:
# locus_table_kenya, locus_table_ghana, locus_table_thailand, locus_table_colombia

# Create the alleles_combined dataframe
alleles_combined <- data.frame(
  Locus = rownames(locus_table_kenya),  # Ensure locus names are extracted properly
  Kenya = as.integer(locus_table_kenya[, "allele"]),
  Ghana = as.integer(locus_table_ghana[, "allele"]),
  Thailand = as.integer(locus_table_thailand[, "allele"]),
  Colombia = as.integer(locus_table_colombia[, "allele"])
)

library(ggplot2)
library(tidyr)

# Reshape the dataframe for easier plotting
alleles_long <- alleles_combined %>%
  pivot_longer(cols = -Locus, names_to = "Population", values_to = "Allele_Count")

# Create a bar plot
ggplot(alleles_long, aes(x = Locus, y = Allele_Count, fill = Population)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Allele Counts by Locus and Population",
       x = "Microsatellite Locus",
       y = "Unique Alleles ") +
  theme_minimal() +
  scale_fill_manual(values = c("Kenya" = "#FF5800",
                                "Ghana" = "#0072B5FF", 
                                "Thailand" = "#EE4C97FF", 
                                "Colombia" = "#20854EFF" ))  # Customize colors as needed




# ------------- Generate a gt table 

library(gt)
library(dplyr)
library(scales)

# Create the gt table
alleles_gt_table <- alleles_combined %>%
  gt() %>%
  tab_header(
    title = "Unique Alleles by Locus"
  ) %>%
  cols_label(
    Locus = "Locus",
    Kenya = "Kenya",
    Ghana = "Ghana",
    Thailand = "Thailand",
    Colombia = "Colombia"
  ) %>%
  fmt_number(
    columns = everything(),
    decimals = 0  # Ensure no decimal places
  )

# Print the gt table
print(alleles_gt_table)

```

## minimum spanning network

https://grunwaldlab.github.io/poppr/reference/poppr.msn.html

```{r}
library("igraph") #install if not already installed

adist <- diss.dist(pfms, percent = TRUE) # calculate dissimilarity matrix
amsn <- poppr.msn(pfms, adist, showplot = TRUE ) # plot default msn

set.seed(500) # set random seed for network shape 

#open 
msn.p <- plot_poppr_msn(pfms, amsn, gadj = 15, layfun = layout_with_kk, vertex.label = NA) # plot
```

```{r}
adist <- diss.dist(pfms, percent = TRUE) # calculate dissimilarity matrix
amsn <- poppr.msn(pfms, adist, showplot = TRUE ) # plot default msn

set.seed(500) # set random seed for network shape 

values = c("Kenya" = "#FF5800", 
                                "Ghana" = "#A1378B", 
                                "Thailand" = "#003A59", 
                                "Colombia" = "#416322FF")

msn.p_ghana <- plot_poppr_msn(pfms, 
                        amsn, 
                        gadj = 5,
                        layfun = layout_as_tree , #layout_with_kk
                        mgl=TRUE,
                        palette = "terrain.colors") # plot

```

```{r}
library(poppr)
library(ggplot2)

# Calculate dissimilarity matrix
adist <- diss.dist(pfms, percent = TRUE)

# Generate the minimum spanning network (MSN)
amsn <- poppr.msn(pfms, adist, showplot = FALSE)

# Set the random seed for reproducibility
set.seed(500)

# Define the custom colors without names
colors <- c("#FF5800",  # Kenya
            "#A1378B",  # Ghana
            "#003A59",  # Thailand
            "#416322FF") # Colombia

# Define population labels corresponding to colors
populations <- c("Kenya", "Ghana", "Thailand", "Colombia")

# Plot the minimum spanning network with custom colors
msn.p_ghana <- plot_poppr_msn(
  pfms, 
  amsn, 
  gadj = 5,
  layfun = layout_as_tree,  # Use tree layout
  mgl = TRUE,
  vertex.color = colors,# Assign the custom colors to vertices
  inds = "reference"
)

# Display the plot
print(msn.p_ghana)

# Add a legend with specific x and y coordinates
legend(x = 0.1,  # X coordinate (10% from the left)
       y = 1.1,  # Y coordinate (90% from the bottom)
       legend = populations, 
       fill = colors, 
       title = "Population", 
       bty = "n",  # No box around the legend
       cex = 0.8)  # Adjust the text size

```

specific labels

```{r}
library(poppr)
library(igraph)  # Make sure to load igraph for layout functions

# Calculate dissimilarity matrix
adist <- diss.dist(pfms, percent = TRUE)

# Generate the minimum spanning network (MSN) and ensure it's a graph object
amsn <- poppr.msn(pfms, adist, showplot = FALSE)

# Set the random seed for reproducibility
set.seed(500)

# Define the custom colors
colors <- c("#FF5800",  # Kenya
            "#A1378B",  # Ghana
            "#003A59",  # Thailand
            "#416322FF") # Colombia

# Define population labels corresponding to colors
populations <- c("Kenya", "Ghana", "Thailand", "Colombia")

# Plot the minimum spanning network with custom colors
msn.p_ghana <- plot_poppr_msn(
  pfms, 
  amsn, 
  gadj = 5,
  layfun = layout_as_tree,  # Use tree layout
  mgl = TRUE,
  vertex.color = colors # Assign the custom colors to vertices
)

# Display the plot
print(msn.p_ghana)


```

## Diagnostic: genotype accumulation curve

> Genotype accumulation curves are useful for determining the minimum number of loci necessary to discriminate between individuals in a population. This function will randomly sample loci without replacement and count the number of multilocus genotypes observed.

```{r}
library(tidyverse)

# Plot the genotype curve
genotype_curve(
  pfms, #import genalex imported file 
  sample = 100,
  #maxloci = 0L,
  quiet = FALSE,
  plot = TRUE,
  drop = FALSE, # TRUE removes monomorphic loci 
  dropna = TRUE # No missing sites
)


#p_gc <- last_plot() # call in last plot 
#p_gc + geom_smooth() + 
 # theme_bw() + 
  #theme(text = element_text(size = 12, family = "serif")) +
  #theme(title = element_text(size = 14)) +
  #ggtitle("Genotype Accumulation Curve for Kenyan Clones", "12 Microsattelite Markers")
 
#`geom_smooth()` using method = 'gam' and formula 'y ~ s(x, bs = "cs")'

```

For our example dataset, it looks like we only need 10 loci to describe the full diversity in our population.

## Adegenet Interactive Server

From: https://adegenet.r-forge.r-project.org/files/PRstats/practical-MVAintro.1.0.pdf

```{r}
#adegenetServer(what = "DAPC")
```

## DPAC Analysis

To start we will be performing descriminant principle component analysis from the [DPAC](https://github.com/thibautjombart/adegenet/raw/master/tutorials/tutorial-dapc.pdf) tutorial.

We will use the same data as before and convert the `genclone` object to a `genid` object for use in adegenet

```{r}
# convert pfms to gen ID 
#pfms.gid <- genclone2genind(pfms)

pfms.gid <- pfms
```

## Find Clusters

specify 2 PCs and 5 Clusters

```{r}
#| eval: false 
clust <- find.clusters(pfms.gid, max.n.clust=10) #requires specification of clusters and values 
clust <- find.clusters(pfms, max.n.clust=10) #requires specification of clusters and values 


```

specify 5 and 5

```{r}

dapc1 <- dapc(pfms, clust$clust) 
```

### Plot dapc with adegenet scatter

```{r}
myCol <- c("red",
           "#827100",
           "forestgreen",
           "blue",
           "purple",
           "green")

# Plot the DAPC scatter plot without automatic labels
scatter(dapc1, 
        posi.da="none",  
        posi.pca="topleft",
        bg="white",
        cstar=0, 
        col=myCol, 
        scree.pca=TRUE,
        cex=2,
        solid=0.7)


```

## FINAL PCA PLOT

```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(ggside)
library(adegenet)
library(ggrepel)  # For improved label placement
library(ggforce)

# Assuming dapc1 contains PCA coordinates and region assignments
coords <- dapc1$ind.coord
sample_ids <- rownames(coords)
region_assignments <- dapc1$grp  # Assuming this is the region assignment

# Combine the extracted data into a data frame
dapc_data <- data.frame(
  SampleID = sample_ids,
  PC1 = coords[, 1],
  PC2 = coords[, 2],
  Region = region_assignments  # Include Region for coloring
)

# Define the specific sample IDs you want to label
reference_ids <- c("3D7", "7G8", "Dd2", "FVO(NIH)", "NF54")

# Create a new column for labels, only showing reference IDs
dapc_data <- dapc_data %>%
  mutate(Label = ifelse(SampleID %in% reference_ids, SampleID, ""))

# Create a separate dataset for ellipses, excluding the reference IDs
dapc_data_no_references <- dapc_data %>%
  filter(!SampleID %in% reference_ids)

# Calculate the percentage of variance explained by each PC
variance_explained <- dapc1$pca.eig
percent_variance_explained <- variance_explained / sum(variance_explained) * 100

# Extract PC1 and PC2 percent variance explained
pc1_variance <- round(percent_variance_explained[1], 2)  # PC1 variance
pc2_variance <- round(percent_variance_explained[2], 2)  # PC2 variance

# Define your custom color scheme

country_colors <- c( 
  "Thailand" = "#003A59", 
  "Cambodia" = "#003A59", 
  "Colombia" = "#416322FF", 
  "Kenya" = "#FF5800",
  "Ghana" = "#A1378B"
)

country_colors <- c( 
                  "Thailand" = "#EE4C97FF", 
                  "Cambodia" = "#EE4C97FF", 
                  "Colombia" = "#20854EFF", 
                  "Ghana" = "#0072B5FF",
                  "Kenya" = "#FF5800"
                 )

# Create the base scatter plot with ggplot2
scatter_plot <- ggplot(dapc_data, aes(x = PC1, y = PC2, fill = Region)) +
  
geom_mark_ellipse(data = dapc_data_no_references, 
                  aes(fill = Region), 
                  alpha = 0.1, 
                  label.buffer = unit(0, "mm"),  # Ensures no buffer around label
                  color = NA,  # Removes the outline of the ellipse
                  expand = unit(2, "mm"),  # Relaxes the contour slightly
                  show.legend = FALSE) +
  
  # Add points for all samples, including reference points, with black outlines
  geom_point(size = 2.5, alpha = 0.7, shape = 21, color = "black") +
  
  # Improved label placement with ggrepel, and white background behind text for reference points
  geom_label_repel(aes(label = Label), 
                   size = 3, 
                   fontface = "bold", 
                   color = "black", 
                   fill = "white",  # White background for labels
                   box.padding = 0.35, 
                   point.padding = 0.5,
                   segment.color = "grey50",
                   max.overlaps = 30) +
  
  # Minimal theme for a clean look
  theme_bw() +  
  labs(
    x = paste("PC1 -", pc1_variance, "% variance explained"),
    y = paste("PC2 -", pc2_variance, "% variance explained"),
    color = "Region"
  ) +
  
  # Apply custom color scheme
  scale_color_manual(values = region_colors) +
  scale_fill_manual(values = region_colors) +
  
  theme(
    legend.position = "right",
    panel.grid.major = element_blank(),  # Remove major gridlines
    panel.grid.minor = element_blank(),  # Remove minor gridlines
    panel.border = element_blank(),      # Remove plot border
    axis.line = element_line(color = "black")  # Add axis lines for clarity
)

# Add marginal histograms using ggside, excluding outlines and adding back horizontal gridlines for y-side
scatter_plot_with_margins <- scatter_plot +
  geom_xsidehistogram(position = "stack") + 
  geom_ysidehistogram(position = "stack") +
  theme(
    legend.position = c(0.65, 0.08),  # Adjust these values to position the legend
    legend.justification = c("left", "bottom")  # Adjusts the reference point for positioning
  )


scatter_plot_with_margins

# ---- save plot

ggsave("figures/PCA.png", 
       scatter_plot_with_margins, 
       height=5, 
       width=5,
       dpi = 500)

```

## Loadings plot

```{r}
set.seed(4)
contrib <- loadingplot(dapc1$var.contr, axis=2,
thres=.07, lab.jitter=1)
```

## unsupervised clusters

```{r}
scatter(dapc1, scree.da=FALSE, bg="white", pch=20,  cell=0, cstar=0, col=myCol, solid=0.4,
        cex=3,clab=0, leg=TRUE, txt.leg=paste("Cluster",1:5))
```

```{r}
scatter(dapc1, ratio.pca=0.3, bg="white", pch=20,  cell=0,
        cstar=0, col=myCol, solid=.4, cex=3, clab=0,
        mstree=TRUE, scree.da=FALSE, posi.pca="bottomright",
        leg=TRUE, txt.leg=paste("Cluster",1:5))
par(xpd=TRUE)
points(dapc1$grp.coord[,1], dapc1$grp.coord[,2], pch=4,
       cex=3, lwd=8, col="black")
points(dapc1$grp.coord[,1], dapc1$grp.coord[,2], pch=4,
       cex=3, lwd=2, col=myCol)

myInset <- function(){
  temp <- dapc1$pca.eig
  temp <- 100* cumsum(temp)/sum(temp) 
  plot(temp, col=rep(c("black","lightgrey"),
               c(dapc1$n.pca,1000)), ylim=c(0,100),
         xlab="PCA axis", ylab="Cumulated variance (%)",
         cex=1, pch=20, type="h", lwd=2)
}


add.scatter(myInset(), posi="bottomright",
            inset=c(-0.03,-0.01), ratio=.28,
            bg=transp("white"))
```

## Loading plot

```{r}
set.seed(4)
#contrib <- loadingplot(dapc.flu$var.contr, axis=2,
#                      thres=.07, lab.jitter=1)
```

```{r}
scatter(dapc1,1,1, col=myCol, bg="white",
        scree.da=FALSE, legend=TRUE, solid=.4)
```

## Population Statistics

```{r}

library(poppr)

poppr_diversity_ci = diversity_ci(
  pfms,
  n = 100L,
  n.boot = 1L,
  ci = 95,
  total = FALSE,
  rarefy = FALSE,
  n.rare = 1,
  plot = TRUE,
  raw = TRUE,
  center = TRUE  
) 

poppr_plot <- last_plot()

ggsave("figures/poppr_ci.svg", poppr_plot, height = 4, width = 6)

poppr_diversity_ci 

```

## Formatted Population Statistics

```{r}
# Specify colors for each population (based on countries)
country_colors <- c( 
                  "Thailand" = "#EE4C97FF", 
                  "Cambodia" = "#EE4C97FF", 
                  "Colombia" = "#20854EFF", 
                  "Ghana" = "#0072B5FF",
                  "Kenya" = "#FF5800"
                 )


# Loop through each metric to create separate plots
for (metric in diversity_metrics) {
  # Subset the data for the current metric
  metric_data <- subset(merged_df, variable == metric)
  
  # Create the plot for the current metric
  p <- ggplot(metric_data, aes(x = Pop)) +
    geom_point(aes(y = value_est, color = Pop), size = 3) +  # Estimated values
    geom_errorbar(aes(ymin = value_obs, ymax = value_obs), width = 0.2) +  # Observed values
    scale_color_manual(values = country_colors) +  # Use the custom population colors
    labs(title = paste("Diversity Index Metric:", metric), x = "Population", y = metric) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), 
          legend.position = "bottom")  # Adjust legend position if needed

  # Store the plot in the list
  plot_list[[metric]] <- p
}

# Display all plots
plot_list[["H"]]     # Plot for H
plot_list[["G"]]     # Plot for G
plot_list[["lambda"]] # Plot for lambda
plot_list[["E.5"]]   # Plot for E.5

```

## Caclulate Genetic Distances 

```{r}

library(writexl)
nei <- nei.dist(pfms, warning = TRUE) 
nei_matrix <- as.matrix(nei)
nei_df <- as.data.frame(nei_matrix)

# save as csv
#write_csv(nei_df, "results/microsattelite_nei.csv")
# save as Excel file
#write_xlsx(nei_df, "results/microsattelite_nei.xlsx")

# Rogers Distance see https://doi.org/10.2307/2407826 and 10.4236/ajps.2016.71022  
rogers <- rogers.dist(pfms)
rogers_matrix <- as.matrix(rogers)
rogers_df <- as.data.frame(rogers_matrix)

# save as csv
write_csv(rogers_df, "results/microsattelite_rogers.csv")
# save as Excel file
write_xlsx(rogers_df, "results/microsattelite_rogers.xlsx")




# other distance metrics
edwards.dist(pfms)
reynolds.dist(pfms)
provesti.dist(pfms)
```

# Cite Packages

```{r}

# Define the packages you want to cite
packages <- c("poppr", "adegenet", "ggplot2", "seqvisr")

# Function to generate citations
generate_citations <- function(pkg) {
  citation_info <- citation(pkg)
  citation_text <- paste(citation_info$authors, "(", citation_info$year, ").", citation_info$title, sep = " ")
  return(citation_text)
}

# Generate citations for all packages
citations <- lapply(packages, generate_citations)

# Print citations (you can customize the output format as needed)
for (citation in citations) {
  cat(citation, "\n")
}
```
